---
title: "Thisisafunnygroupname's Project Report"
author: "Richard Zhou, Adam Rui, Jonathan Darius, Ojasvi Godha, Ryan Huang, Isaac Kang"
output: 
  pdf_document:
    toc: true
---

\newpage

**[DELETE ALL TEXT IN BRACKETS AND TEMPLATE COMMENTS IN CODE WHEN FINISHED]**

# Introduction

[Write a quick introduction]

# Project Description

[Write about the project, our project objectives, and the questions we seek to answer]

Through this data analysis, we aim to answer the 5 following questions:

1.  Have flight delays improved over time overall? 

    -   What about with individual airlines?\

2.  Do busy destinations tend to have more or less delays?\

3.  Is the weather correlated with flight delays?

    -   How has this changed over time?\

4.  Is the time of the year correlated between flight delays (holidays or rainy season)?\

5.  Which airlines have the least delays?

    -   How has this changed over time?

\newpage

# Research Questions

## [REPLACE WITH QUESTION #1]

### Data Exploration and Visualization

```{r}
# reuse/refine the plot made in the proposal
```

[Discuss the visualization. What are some important takeaways? What could we possibly find interesting insights in judging from the plot? Any possible reasons for these insights? Talk about how your visualization leads to your analysis]

### Data Analysis/Modeling/Predictions

```{r}
# code for testing your hypotheses/models

# DON'T FORGET TO CHECK NECESSARY ASSUMPTIONS FOR PERFORMING ANALYSES
# there are plenty of premade functions to test assumptions, just search them up
```

[Discuss your results. Don't forget that no results is still an important conclusion, with plenty to discuss! What are some important takeaways? Any possible explanations for these takeaways? How can we apply this new found knowledge?]

## [Do busy destinations tend to have more or less delays?]

### Data Exploration and Visualization

```{r}
library("tidyverse")
library("nycflights13")
library("nycflights23")
library("dplyr")

flights_combined <- bind_rows(nycflights13::flights %>% mutate(year = 2013), nycflights23::flights %>% mutate(year = 2023))

flights_clean <- flights_combined %>%
  filter(!is.na(dep_delay), !is.na(arr_delay)) %>%
  left_join(nycflights13::airlines, by = "carrier") %>%
  left_join(nycflights13::airports, by = c("dest" = "faa"))

destination_stats <- flights_clean |>
  filter(!is.na(name.y)) |>
  group_by(dest, name.y) |>
  summarise(
    total_flights = n(),
    avg_delay = mean(dep_delay + arr_delay),
    .groups = "drop"
  ) |>
  mutate(
    busyness = total_flights / sum(total_flights),
    busyness_rank = dense_rank(busyness)
  ) %>%
  arrange(busyness_rank)

#_____________________ my stuff starts here the above is the data cleaning stuff_______________________
library("ggrepel") #for the floating text labels

important_airports <- destination_stats |>
  arrange(desc(avg_delay)) |>
  slice(c(1:5, (n()-4):n())) |>
  bind_rows(
    destination_stats |>
      arrange(desc(busyness)) |> 
      slice(1:5)  # 5 busiest
  ) |>
  distinct(dest, .keep_all = TRUE)

#for the correlation and p value
cor_test <- cor.test(destination_stats$busyness, destination_stats$avg_delay)
correlation <- cor_test$estimate
p_value <- cor_test$p.value

ggplot(destination_stats, aes(x = busyness, y = avg_delay)) +
  geom_point(aes(size = total_flights, color = avg_delay), alpha = 0.5) +
  
  #linear fit line
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  
  #floating text for impotant airports
  geom_text_repel(
    data = important_airports,
    aes(label = paste(dest, name.y)), 
    size = 3,
    box.padding = 0.5
  ) +
  
  #add colors to visualise delay better
  scale_color_gradient2(
    low = "green", mid = "blue", high = "red", 
    midpoint = median(destination_stats$avg_delay)
  ) +
  labs(
    x = "Proportion of Total Flights (Busyness)",
    y = "Average Delay (minutes)",
    title = "Flight Delays vs. Destination Busyness",
    subtitle = sprintf(
      "Correlation: %.2f (p = %.3f)", 
      correlation, 
      p_value
    ),
    size = "Total Flights",
    color = "Avg Delay"
  )
```

[Discuss the visualization. What are some important takeaways? What could we possibly find interesting insights in judging from the plot? Any possible reasons for these insights? Talk about how your visualization leads to your analysis]

### Data Analysis/Modeling/Predictions

```{r}
library(lmtest)
model <- lm(avg_delay ~ busyness, data = destination_stats)
bptest(model)  # p > 0.05 = homoscedastic

shapiro.test(residuals(model))

#accounting for heteroscedasticity (obust standard error)
library(sandwich)


#accounting for normality (np regression)
library(mgcv)
model_gam <- gam(avg_delay ~ s(busyness), data = destination_stats)
summary(model_gam)
```

[Discuss your results. Don't forget that no results is still an important conclusion, with plenty to discuss! What are some important takeaways? Any possible explanations for these takeaways? How can we apply this new found knowledge?]

## [REPLACE WITH QUESTION #3]

### Data Exploration and Visualization

```{r}
# reuse/refine the plot made in the proposal
```

[Discuss the visualization. What are some important takeaways? What could we possibly find interesting insights in judging from the plot? Any possible reasons for these insights? Talk about how your visualization leads to your analysis]

### Data Analysis/Modeling/Predictions

```{r}
# code for testing your hypotheses/models  

# DON'T FORGET TO CHECK NECESSARY ASSUMPTIONS FOR PERFORMING ANALYSES # there are plenty of premade functions to test assumptions, just search them up
```

[Discuss your results. Don't forget that no results is still an important conclusion, with plenty to discuss! What are some important takeaways? Any possible explanations for these takeaways? How can we apply this new found knowledge?]

## [Question #4: Does time of year affect flight delays?]

### Data Exploration and Visualization

```{r}
flights_clean %>%
  mutate(month = month(time_hour, label = TRUE)) %>% #gets the month from time_hour
  group_by(month, year) %>%
  summarise(avg_dep_delay = mean(dep_delay), .groups = 'drop') %>% #computes the average departure delay for that month
  ggplot(aes(x = month, y = avg_dep_delay, group = year, color = factor(year))) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 4) +
  labs(title = "Seasonal Pattern of Departure Delays", x = "Month", y = "Avg Departure Delay (mins)", color = "Year") +
  theme_bw()
```

- This line chart shows how departure delays vary across months for both years. Peaks in certain months could point to holiday seasons, weather events, or seasonal congestion affecting flight performance.

### Data Analysis/Modeling/Predictions

```{r}
# Prepare data with numeric month
flights_seasonal <- flights_clean %>%
  mutate(month = month(time_hour, label = FALSE))

# Run one-way ANOVA
anova_model1 <- aov(dep_delay ~ as.factor(month), data = flights_seasonal)
summary(anova_model1)

# Run two-way ANOVA
anova_model <- aov(dep_delay ~ as.factor(month)*as.factor(year), data = flights_seasonal)
summary(anova_model)

# Linear model for two-way ANOVA to calculate adjusted R-squared
lm1 <- lm(dep_delay ~ as.factor(month)*as.factor(year), data = flights_seasonal)
summary(lm1)$adj.r.squared

# constant variance: Levene's test for homogeneity of variance across months
leveneTest(dep_delay ~ as.factor(month), data = flights_seasonal)

# Normality, large sample size sensitive to tests, use graph

# Durbin-Watson test for autocorrelation/seasonal trend.
library(lmtest)  
dwtest(anova_model)     
```

- This visualization shows the seasonal pattern of average departure delays across different months and years. Each line represents a year, with months on the x-axis and the average departure delay (in minutes) on the y-axis.


## [REPLACE WITH QUESTION #5]

### Data Exploration and Visualization

```{r}
# reuse/refine the plot made in the proposal
```

[Discuss the visualization. What are some important takeaways? What could we possibly find interesting insights in judging from the plot? Any possible reasons for these insights? Talk about how your visualization leads to your analysis]

### Data Analysis/Modeling/Predictions

```{r}
# code for testing your hypotheses/models  

# DON'T FORGET TO CHECK NECESSARY ASSUMPTIONS FOR PERFORMING ANALYSES # there are plenty of premade functions to test assumptions, just search them up
```

[Discuss your results. Don't forget that no results is still an important conclusion, with plenty to discuss! What are some important takeaways? Any possible explanations for these takeaways? How can we apply this new found knowledge?]

\newpage

# Conclusions

1.  **Have flight delays improved over time overall?**

    -   **What about with individual airlines?**

[Write a quick paragraph recapping conclusions made from your analysis]

2.  **Do busy destinations tend to have more or less delays?**

[Write a quick paragraph recapping conclusions made from your analysis] i will do this tmrw i;m so sleepy

3.  **Is the weather correlated with flight delays?**

    -   **How has this changed over time?**

[Write a quick paragraph recapping conclusions made from your analysis]

4.  **Is the time of the year correlated between flight delays (holidays or rainy season)?**

[Write a quick paragraph recapping conclusions made from your analysis]

5.  **Which airlines have the least delays?**

    -   How has this changed over time?

[Write a quick paragraph recapping conclusions made from your analysis]

\newpage

# Authors' Contributions

| Author           | Contributions |
|------------------|---------------|
| Richard Zhou     |               |
| Adam Rui         |               |
| Jonathan Darius  |               |
| Ojasvi Godha     |               |
| Ryan Huang       | Question 2    |
| Isaac Kang       |               |
