---
title: "Project Proposal"
authors: "Richard Zhou, Adam Rui, Jonathan Darius, Ojasvi Godha, Ryan Huang, Isaac Kang"
output: html_document
---

## Installation and Packages

```{r}
# install.packages("tidyverse")
# install.packages("nycflights13")
# install.packages("nycflights23")

library("tidyverse")
library("nycflights13")
library("nycflights23")
```

## Objectives:

Our overall objective of this project is to analyze and compare flight data from 2013 and 2023 using the nycflights13 and nycflights23 packages. We aim to identify patterns, trends, and key factors affecting flight performance, delays, and operations over time. Our research aims to answer the following questions:

1.   Have flight delays improved or gotten worse between 2013 and 2023?

2.  To expand on Question #1, have individual airlines gotten better or worse with delays over time?

3.  How does weather impact flight performance, and has it changed over time?

4.  Which airports have seen the largest increase or decrease in traffic?

5.  Have there been any big changes in popular destinations from NYC between 2013 and 2023?

6.  Is there a seasonal pattern to flight delays? (Such as holidays)

7.  What are the most reliable airlines, and has this changed over time?

## Datasets:

[Include your exploratory data analysis and some visualization! Make sure the data is reasonably clean and contains enough information to answer your questions. Which variables in the table are most relevant to your questions? Do you plan to acquire additional datasets?]\
\
We plan on using the packages nycflights13 and nycflights23, using the datasets provided in both. Specifically, we will be focusing on flights, airlines, airports, and weather from both packages.

```{r}
# listing dimensions of datasets
dimensions <- tibble(name = character(), rows = integer(), cols = integer())

dimensions <- add_row(dimensions, name = "flights_2013", rows = nrow(nycflights13::flights), cols = ncol(nycflights13::flights))
dimensions <- add_row(dimensions, name = "flights_2023", rows = nrow(nycflights23::flights), cols = ncol(nycflights23::flights))

dimensions <- add_row(dimensions, name = "airlines_2013", rows = nrow(nycflights13::airlines), cols = ncol(nycflights13::airlines))
dimensions <- add_row(dimensions, name = "airlines_2023", rows = nrow(nycflights23::airlines), cols = ncol(nycflights23::airlines))

dimensions <- add_row(dimensions, name = "airports_2013", rows = nrow(nycflights13::airports), cols = ncol(nycflights13::airports))
dimensions <- add_row(dimensions, name = "airports_2023", rows = nrow(nycflights23::airports), cols = ncol(nycflights23::airports))

dimensions <- add_row(dimensions, name = "weather_2013", rows = nrow(nycflights13::weather), cols = ncol(nycflights13::weather))
dimensions <- add_row(dimensions, name = "weather_2023", rows = nrow(nycflights23::weather), cols = ncol(nycflights23::weather))

print(dimensions)
```

## Our Plan:

[What do you plan to do with the data to answer your questions? For example, what assumptions will you make? What plots will you generate? What models will you explore? What evaluation approaches will you use? Some preliminary data analysis and modeling results will be a plus, but not a must.]

## Alternative Strategies/Backup Plans:

[What if your ideas won’t work out? Include a section called “Alternative Strategies/Backup Plans”, list 2-3 other ideas you have brainstormed, each with a couple bullet points of details.]
